\documentclass[11pt]{article}
\usepackage{vinotheca}
\usepackage{pdflscape}

\begin{document}

\vinothecaTitle{TasteRank Explorer}{Data Appendix}{Source Data, Tables, and Reference Materials}
\vinothecaAuthor{Jure Skarabot}{May 2026}
\vinothecaCompanions{Summary \textperiodcentered{} Technical Appendix \textperiodcentered{} Methods Primer \textperiodcentered{} Data Appendix \textperiodcentered{} Grape Reference}

\section*{Table of Contents}
\begin{itemize}[leftmargin=2em,itemsep=0.1em,topsep=0.1em]
\item D.1 Sensory Dimension Key
\item D.2 Community Summary
\item D.3 Complete Variety Data (Ranks 1--101)
\end{itemize}

\section*{D.1 Sensory Dimension Key}
The following abbreviations are used in the data tables. All scores are integers on a 0--5 scale representing canonical varietal typicity across thirteen sensory dimensions drawn from established wine-tasting methodology (UC Davis, Peynaud, and related structured tasting frameworks).

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
Abbr.\ & Dimension & Description \\
\midrule
Col & Color Depth         & Intensity of pigmentation (pale to opaque) \\
Aro & Aromatic Intensity  & Strength of nose from low to pronounced \\
Flo & Floral              & Degree of floral aromatic contribution \\
Frt & Fruit Ripeness      & Ripe / dried fruit vs.\ underripe / green fruit \\
Hrb & Herbal / Earthy     & Green herb, earth, mineral notes \\
Spk & Spice / Oak         & Oak-derived and spice character \\
Acd & Acidity             & Perceived acidity from low to high \\
Tan & Tannin              & Tannin level (reds); 0 for most whites \\
Bod & Body                & Weight and texture on palate \\
Alc & Alcohol             & Perceived alcohol warmth \\
Fla & Flavor Intensity    & Palate flavour concentration \\
Fin & Finish              & Length and persistence of aftertaste \\
Cpx & Complexity          & Layeredness and evolution of character \\
\bottomrule
\end{tabular}
\end{center}

\section*{D.2 Community Summary}
Six communities were detected by the Clauset--Newman--Moore greedy modularity optimisation algorithm. The table below summarises each community's character, size, colour composition, and hub varieties.

\begin{center}
\begin{tabular}{@{}clrlp{6.5cm}@{}}
\toprule
ID & Character & Size & R / W & Hub Varieties \\
\midrule
C0 & Full-bodied Mediterranean Reds   & 31 & 31 / 0 & Sagrantino, Nero d'Avola, Lagrein, Negroamaro, Plavac Mali \\
C1 & Light / Aromatic Cross-boundary  & 21 & 4 / 17 & Assyrtiko, Falanghina, Albari\~no, Riesling, Kadarka \\
C2 & Mid-weight Structured Reds       & 18 & 18 / 0 & Limniona, St.\ Laurent, Zweigelt, Blaufr\"ankisch, Sangiovese \\
C3 & Rich Textural Whites             & 13 & 0 / 13 & Marsanne, Fiano, Godello, Pinot Gris, Chardonnay \\
C4 & Mineral / Crisp Whites           & 10 & 0 / 10 & Greco, Friulano, Verdicchio, Verdejo, Vermentino \\
C5 & Lean Neutral \& Aromatic Whites  &  8 & 0 / 8  & Gew\"urztraminer, Petit Manseng, Muscat Blanc, Malvasia \\
\bottomrule
\end{tabular}
\end{center}

\begin{landscape}
\section*{D.3 Complete Variety Data (Ranks 1--101)}
All 101 grape varieties sorted by TasteRank. Columns show rank, variety name, type (Red / White), community assignment, TasteRank score, and the thirteen sensory profile dimensions.

\renewcommand{\arraystretch}{1.05}
\small
\begin{longtable}{@{}r l c c r *{13}{c}@{}}
\toprule
\# & Variety & Type & Com & TR & Col & Aro & Flo & Frt & Hrb & Spk & Acd & Tan & Bod & Alc & Fla & Fin & Cpx \\
\midrule
\endfirsthead
\toprule
\# & Variety & Type & Com & TR & Col & Aro & Flo & Frt & Hrb & Spk & Acd & Tan & Bod & Alc & Fla & Fin & Cpx \\
\midrule
\endhead
\bottomrule
\endlastfoot
\input{data_rows.tex}
\end{longtable}
\end{landscape}

\section*{References}

\begin{itemize}[leftmargin=2em,itemsep=0.2em]
\item d'Agata, I. (2014). \emph{Native Wine Grapes of Italy}. University of California Press.
\item Galet, P. (2000). \emph{Dictionnaire encyclop\'edique des c\'epages}. Hachette, Paris.
\item Peynaud, \'E. (1980). \emph{Le Go\^ut du Vin}. Dunod, Paris.
\item Robinson, J., Harding, J. \& Vouillamoz, J. (2012). \emph{Wine Grapes: A Complete Guide to 1{,}368 Vine Varieties}. Ecco / HarperCollins.
\end{itemize}

\end{document}
